---
title: "Classical Models Practical 1"
author: "Isaac Ambrogetti and Andy Muchyo Nkunzimana"
date: "2026-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Solve ordinary differentiation equations numerically

### Malthus model
```{r}
if(!require("deSolve")) install.packages("deSolve")
library("deSolve")
```

```{r}
malthus <- function(t, N0, parms){
  r <- parms["r"]
  dN <- r * N0
  return(list(dN))
}

N0 <- 8
t <- seq(0, 100, by=1)
r <- c(0.1, 0, -0.1)

out <- list()

for (i in seq_along(r)){
  out[[i]] <- ode(N0, t, malthus, parms = c(r = r[i]))
}
```

```{r}
plot(NA, xlim = c(0, 100), ylim = c(1, max(out[[1]][,"1"])),
     log = "y", xlab = "Time", ylab = "N(t) [log scale]", main = "Malthus Model")

for (i in seq_along(r)) {
  out[[i]] <- ode(N0, t, malthus, parms = c(r = r[i]))
  lines(out[[i]][, "time"], out[[i]][, "1"], col = i)
}

legend("topright", legend = paste("r =", r), col = seq_along(r), lty = 1)
```


### Verhulst model

Similar to the Malthus model but it uses the parameter K (carrying capacity) = r / alpha. 
It is the value in population size at which the per capita growth rate is equal to zero
and, therefore, the population stops growing. (population cap)

Alpha is the coefficient of intraspecific competition: slope of the curve in absolute 
value (-alpha is the slope).



```{r}
verhulst <- function(t, N0, parms){
  r <- parms["r"]
  K <- parms["k"]
  dN <- r * N0 * (1 - N0/K)
  return(list(dN))
}

N0 <- 8
t <- seq(0, 100, by=1)
r <- 0.1
k <- c(2, 40, 100)

out <- list()

for (i in seq_along(k)){
  out[[i]] <- ode(N0, t, verhulst, parms = c(r = r, k = k[i]))
}
```

```{r}
plot(NA, xlim = c(0, 100), ylim = c(1, 100),
     log = "y", xlab = "Time", ylab = "N(t) [log scale]", main = "Verhulst Model")

for (i in seq_along(k)) {
  lines(out[[i]][, "time"], out[[i]][, "1"], col = i)
}

legend("topright", legend = paste("k =", k), col = seq_along(k), lty = 1)
```





