---
title: "Classical Models Practical 1"
author: "Isaac Ambrogetti and Andy Muchyo Nkunzimana"
date: "2026-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Solve ordinary differentiation equations numerically

### Malthus model
```{r}
if(!require("deSolve")) install.packages("deSolve")
library("deSolve")
```

```{r}
malthus <- function(t, N0, parms){
  r <- parms["r"]
  dN <- r * N0
  return(list(dN))
}

N0 <- 8
t <- seq(0, 100, by=1)
r <- c(0.1, 0, -0.1)

out <- list()

for (i in seq_along(r)){
  out[[i]] <- ode(N0, t, malthus, parms = c(r = r[i]))
}
```

```{r}
plot(NA, xlim = c(0, 100), ylim = c(1, max(out[[1]][,"1"])),
     log = "y", xlab = "Time", ylab = "N(t) [log scale]", main = "Malthus Model")

for (i in seq_along(r)) {
  out[[i]] <- ode(N0, t, malthus, parms = c(r = r[i]))
  lines(out[[i]][, "time"], out[[i]][, "1"], col = i)
}

legend("topright", legend = paste("r =", r), col = seq_along(r), lty = 1)
```


### Verhulst model

Similar to the Malthus model but it uses the parameter K (carrying capacity) = r / alpha. 
It is the value in population size at which the per capita growth rate is equal to zero
and, therefore, the population stops growing. (population cap)

Alpha is the coefficient of intraspecific competition: slope of the curve in absolute 
value (-alpha is the slope).



```{r}
verhulst <- function(t, N0, parms){
  r <- parms["r"]
  K <- parms["k"]
  dN <- r * N0 * (1 - N0/K)
  return(list(dN))
}

N0 <- 8
t <- seq(0, 100, by=1)
r <- 0.1
k <- c(2, 40, 100)

out <- list()

for (i in seq_along(k)){
  out[[i]] <- ode(N0, t, verhulst, parms = c(r = r, k = k[i]))
}
```

```{r}
plot(NA, xlim = c(0, 100), ylim = c(1, 100),
     log = "y", xlab = "Time", ylab = "N(t) [log scale]", main = "Verhulst Model")

for (i in seq_along(k)) {
  lines(out[[i]][, "time"], out[[i]][, "1"], col = i)
}

legend("topright", legend = paste("k =", k), col = seq_along(k), lty = 1)
```

### Original Lotka-Volterra model

```{r}
LV <- function(t, pop, parms){
  r <- parms["r"]
  gamma <- parms["gamma"]
  m <- parms["m"]
  beta <- parms["beta"]
  alpha <- parms["alpha"]
  
  dN <- (r - alpha*pop["N"] - gamma*pop["P"])*pop["N"]
  dP <- (-m + beta*pop["N"]) * pop["P"]
  return(list(c(dN, dP)))
}

t <- seq(1, 200, by=1)
pop <- c(N = 10, P = 8)
r <- 1
gamma <- 0.2
m <- 0.2
beta <- 0.1
alpha <- 0.2
scale <- 0.1

out <- ode(pop, t, LV, parms = c(r = r, gamma = gamma, m = m, beta = beta, alpha = alpha))
df.out <- as.data.frame(out)
```

```{r}
plot(NULL,xlim=c(0, 8), ylim=c(0, 10), xlab="N", ylab="P")

abline(a = r/gamma, b = - (alpha/gamma), col="blue")
abline(v = m/beta, col = "red")

N <- seq(0.1, 5, length.out=30)
P <- seq(0.1, 8, length.out=30)

for (p in P){
  for (n in N){
    n1 <- n + (r - alpha*n - gamma*p)*n * scale
    p1 <- p + (-m + beta*n)*p * scale
    arrows(n, p, n1, p1, length = 0.05, angle = 30, code = 2, col = "lightgreen")
  }
}
  
lines(df.out$N, df.out$P, type="l", col="brown")
```



### Original Kermack-McKendrick’s SIR epidemic model

```{r}
SIR <- function(time, state, params){
  S <- state["S"] 
  I <- state["I"] 
  alpha <- params["alpha"] 
  beta <- params["beta"] 
  K <- params["K"] 
  r <- params["r"] 
  
  dS <- (r * (1 - (S/K))) * S - (beta * S * I)
  dI <- (beta * S * I) - (alpha * I)
  
  return(list(c(dS, dI)))
}

t <- seq(1, 50, by=1)
pop <- c(S = 2, I = 0.1)     # S = susceptible population, I = infected population

params <- c(
  alpha=0.2,  # rate at which infected host leaves this class (dies or recovers)
  beta=0.5,   # rate at which susceptible becomes infected (increase of infected pop)
  K=2,   # carrying capacity, max pop size of a specific population, given by r/alpha, intrinsic growth rate divided by the coeff of intraspecific competition"
  r=1    # intrinsic growth rate of the population
)

scale <- 0.1  # scale for vector size

out <- ode(pop, t, SIR, parms = params)
df.out <- as.data.frame(out)
```

```{r}
plot(df.out$I~df.out$S, xlab="S", ylab="I")

S <- seq(0.1, 5, length.out=30)
I <- seq(0.1, 8, length.out=30)

abline(v = alpha/beta, col="blue")
abline(a = 1/beta, b = - S/(beta*k), col = "red")
```

```{r}
epid_thresh <- params["alpha"] / params["beta"]
# --- Vector field grid ---
S_seq <- seq(0.01, 2.5, length.out = 20)
I_seq <- seq(0.01, 2.5, length.out = 20)
grid  <- expand.grid(S = S_seq, I = I_seq)

# Compute derivatives at each grid point
dS_vec <- numeric(nrow(grid))
dI_vec <- numeric(nrow(grid))
for(k in seq_len(nrow(grid))){  
  state_k <- c(S = grid$S[k], I = grid$I[k])  
  derivs <- SIR(0, state_k, params)[[1]]  
  dS_vec[k] <- derivs[1]  
  dI_vec[k] <- derivs[2]
  }
# Normalize arrow lengths for readability
magnitude <- sqrt(dS_vec^2 + dI_vec^2)
scale <- 0.08  # adjust this to make arrows longer/shorter
dS_norm <- scale * dS_vec / magnitude
dI_norm <- scale * dI_vec / magnitude

plot(
  df.out$I~df.out$S,
  xlab="S",
  ylab="I",
  #xlim=c(0, 1.5),
  #ylim=c(0,5)
  )

arrows(  
  x0 = grid$S,  
  y0 = grid$I,
  x1 = grid$S + dS_norm,
  y1 = grid$I + dI_norm,  
  length = 0.04, # arrowhead size 
  col = adjustcolor("steelblue", alpha.f = 0.5),  
  lwd = 0.8
  )
abline(v = epid_thresh, col = "blue")
abline(a = 1/0.5, b = (-1/(2*0.5)), col = "red")
legend(
  "topright",  
  legend = c("Trajectory", "S isocline threshold", "I isocline"),
  col = c("black", "blue", "red"),
  lty = c(1, 2, 2),
  lwd    = 2
  )
```



